---
title: "TP 1"
format: html
editor: visual
---

---
title: "Statistical Models Exercise Answers"
subtitle: "Institute of Technology of Cambodia — I4STM"
author: "SOM Chann Reaksmey, NGEN Tina, NY Chantharith, TIENG Sopanha, YENG Kheang"
description: "WRITING THIS TP WITH LOVE NOT FAKE LIKE CHHAYTHEAN"
format: html
editor: visual
---

# SECTION 1: DATA VISUALIZATION

## Task 1: Galton Heredity

```{r}
library(HistData)
data(Galton)
Galton <- data.frame(Galton)
```

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
```

### (a). Contingency table between children and parents’ height

```{r}
head(Galton)
str(Galton)
summary(Galton)
```

```{r}
child_breaks <- seq(60, 75, by = 0.5)
parent_breaks <- seq(63, 72, by = 0.5)
```

```{r}
child_midpoints <- (child_breaks[-length(child_breaks)] + child_breaks[-1]) / 2
parent_midpoints <- (parent_breaks[-length(parent_breaks)] + parent_breaks[-1]) / 2

Galton$child_group <- cut(Galton$child, breaks = child_breaks, labels = round(child_midpoints, 2))
Galton$parent_group <- cut(Galton$parent, breaks = parent_breaks, labels = round(parent_midpoints, 2))
# Create contigency table
table <- table(Galton$parent_group, Galton$child_group)
table
```

### (b) Scatter plot with regression line

```{r}
# Fit linear regression model
model <- lm(child ~ parent, data = Galton)

# Get regression coefficients
slope <- coef(model)[2]
intercept <- coef(model)[1]
```

```{r}
# Create scatter plot with regression line
plot(Galton$parent, Galton$child,
     main = "Height of Children VS. Average Height of Parents",
     xlab = "Average Parent Height (inches)",
     ylab = "Child Height (inches)",
     pch = 16,
     col = rgb(0, 0, 0, 0.5), # Semi-transparent black points
     cex = 0.8
     )

# Add regression line
abline(model, col = "red", lwd = 2)

# Add equation and R-squared to the plot
r_squared <- summary(model)$r.squared
equation <- paste("y =", round(intercept, 3), "+", round(slope, 3), "* x")
r2_text <- paste("R^2 =", round(r_squared, 4))

legend ("topleft", legend = c(equation, r2_text),
        col = c("red", "black"), lty = c(1, NA),
        bty = "n", cex = 0.9
        )
```

```{r}
# Print model summary
"=== Regression Summary ===\n"
summary(model)

"\n=== Key Statistics ===\n"
cat("Number of observations: ", nrow(Galton), "\n")
cat("Slope: ", round(slope, 4), "\n")
cat("Intercept: ", round(intercept, 4), "\n")
cat("R-squared: ", round(r_squared, 4), "\n")
cat("Correlation: ", round(cor(Galton$parent, Galton$child), 4), "\n")
```

## Task 2: Munich Rent Index (1999)

```{r}
#install.packages("gamlss.data") #for the first time you need to install the package
library(gamlss.data)
data(rent99)
rent99<-data.frame(rent99)
head(rent99)
```

```{r}
# Structure of the data
library(dplyr)
glimpse(rent99)
```

### (a) Histograms and Kernel Density Plots

```{r}
# Load and inspect dataset
head(rent99)
summary(rent99)
```

```{r}
par(mfrow=c(2,2))
hist(rent99$rent, breaks = 30, freq = FALSE, prob=TRUE, main="Net Rent", xlab="Net rent (€)", col="lightblue")
lines(density(rent99$rent), col="red", lwd=2)

hist(rent99$rentsqm, breaks = 30, freq = FALSE, prob=TRUE, main="Net Rent per sqm", xlab="€", col="lightgreen")
lines(density(rent99$rentsqm), col="red", lwd=2)

hist(rent99$area, breaks = 30, freq = FALSE, prob=TRUE, main="Area (sqm)", xlab="sqm", col="lightyellow")
lines(density(rent99$area), col="red", lwd=2)

hist(rent99$yearc, breaks = 30, freq = FALSE, prob=TRUE, main="Year of Construction", xlab="Year", col="darkblue")
lines(density(rent99$yearc), col="red", lwd=2)

```

### (b) Scatter Plots

```{r}
par(mfrow=c(2,2))
plot(rent99$area, rent99$rent, main="Net Rent vs Area", xlab="Area (sqm)", ylab="Net Rent (€)", col="blue")
plot(rent99$area, rent99$rentsqm, main="Rent per sqm vs Area", xlab="Area (sqm)", ylab="Rent/sqm (€)", col="darkgreen")
plot(rent99$yearc, rent99$rent, main="Net Rent vs Year", xlab="Year", ylab="Net Rent (€)", col="purple")
plot(rent99$yearc, rent99$rentsqm, main="Rent per sqm vs Year", xlab="Year", ylab="Rent/sqm (€)", col="orange")
```

### (c) Average and SD Plot

```{r}
library(ggplot2)
# Compute average values per cluster
rent_area <- rent99 %>%
  group_by(area) %>%
  summarise(
    mean_rent = mean(rent, na.rm = TRUE),
    sd_rent = sd(rent, na.rm = TRUE),
    mean_rentsqm = mean(rentsqm, na.rm = TRUE),
    sd_rentsqm = sd(rentsqm, na.rm = TRUE)
  )
rent_yearc <- rent99 %>%
  group_by(yearc) %>%
  summarise(
    mean_rent = mean(rent, na.rm = TRUE),
    sd_rent = sd(rent, na.rm = TRUE),
    mean_rentsqm = mean(rentsqm, na.rm = TRUE),
    sd_rentsqm = sd(rentsqm, na.rm = TRUE)
  )
```

```{r}
ggplot(rent_area, aes(x = area, y = mean_rent)) +
  geom_point(shape = 21, fill = "black", size = 2) +
  geom_errorbar(aes(ymin = mean_rent - sd_rent, ymax = mean_rent + sd_rent),
                width = 0.5, color = "red") +
  labs(title = "Average net rent vs. area",
       x = "area in sqm", y = "average net rent") +
  theme_bw()

ggplot(rent_area, aes(x = area, y = mean_rentsqm)) +
  geom_point(shape = 21, fill = "black", size = 2) +
  geom_errorbar(aes(ymin = mean_rentsqm - sd_rentsqm, ymax = mean_rentsqm + sd_rentsqm),
                width = 0.5, color = "red") +
  labs(title = "Average net rent per sqm vs. area",
       x = "area in sqm", y = "average net rent per sqm") +
  theme_bw()

ggplot(rent_yearc, aes(x = yearc, y = mean_rent)) +
  geom_point(shape = 21, fill = "black", size = 2) +
  geom_errorbar(aes(ymin = mean_rent - sd_rent, ymax = mean_rent + sd_rent),
                width = 0.5, color = "red") +
  labs(title = "Average net rent vs. year of construction",
       x = "year of construction", y = "average net rent") +
  theme_bw()

ggplot(rent_yearc, aes(x = yearc, y = mean_rentsqm)) +
  geom_point(shape = 21, fill = "black", size = 2) +
  geom_errorbar(aes(ymin = mean_rentsqm - sd_rentsqm, ymax = mean_rentsqm + sd_rentsqm),
                width = 0.5, color = "red") +
  labs(title = "Average net rent per sqm vs. year of construction",
       x = "year of construction", y = "average net rent per sqm") +
  theme_bw()
```

### (d) Boxplot and Density by Location

```{r}
plot(rent99$location, rent99$rentsqm,
     xlab = "location", ylab = "net rent per sqm")

# smooth density estimator
# label location categories
rent99$Location <- factor(rent99$location, 
                          levels = c(1, 2, 3), 
                          labels = c("average", "good", "top"))
```

```{r}
net_rent_average <- rent99$rentsqm[rent99$Location == "average"]
net_rent_good <- rent99$rentsqm[rent99$Location == "good"]
net_rent_top <- rent99$rentsqm[rent99$Location == "top"]


plot(density(net_rent_average), 
     main = "",
     xlab = "net rent per sqm",
     ylab = "estimated density",
     xlim = c(0, 18),
     ylim = c(0, 0.16),
     lwd = 2,
     col = "black",
     lty = 1)
lines(density(net_rent_good), lwd = 2, col = "blue", lty =2)
lines(density(net_rent_top), lwd = 2, col = "red", lty =3 )
legend("topright", 
       legend = c("average", "good", "top"),
       lty = c(1, 2, 3),
       col = c("black", "blue", "red"),
       lwd = 2,
       bty = "n")
```

## Task 3: Fuel Consumption

```{r}
#install.packages("alr4") #for the first time you need to install the package
library(alr4)
data(fuel2001)
fuel2001<-data.frame(fuel2001)
head(fuel2001)
```

### (a) Create new variables

```{r}
fuel2001$Fuel <- 1000 * fuel2001$FuelC / fuel2001$Pop
fuel2001$Dlic <- 1000 * fuel2001$Drivers / fuel2001$Pop
fuel2001$logMiles <- log(fuel2001$Miles)
```

```{r}
summary(fuel2001)
head(fuel2001)
```

### (b) Summary and Correlation Plot

```{r}
y <- fuel2001$FuelC
# Examine the response variable
summary(fuel2001$FuelC)
# Visualize distribution of fuel comsumption across all states
hist(fuel2001$FuelC,
     main = "Distribution of Fuel Comsumption Across All States",
     xlab = "Fuel Comsumption (thousands of gallons)",
     col = "lightblue",
     breaks = 15
     )
```

```{r}
install.packages("psych")
library(psych)
describe(fuel2001)

# Drivers
hist(fuel2001$Drivers,
     main = "Distribution of Licensed Drivers",
     xlab = "Number of Drivers",
     col = "lightblue",
     breaks = 20
     )
boxplot(fuel2001$Drivers,
        main = "Boxplot: Licensed Drivers",
        ylab = "Number of Drivers",
        col = "lightblue",
        horizontal = FALSE
        )
# FuelC
hist(fuel2001$FuelC,
     main = "Distribution of Fuel Consumption",
     xlab = "gas sold",
     col = "lightblue",
     breaks = 20
)
boxplot(fuel2001$FuelC,
        main = "Boxplot: Fuel Consumption",
        ylab = "gas sold",
        col = "lightblue"
        )
# Income
hist(fuel2001$Income,
     main = "Distribution of Income",
     xlab = "Income per person",
     col = "lightblue",
     breaks = 20
    )
boxplot(fuel2001$Income,
        main = "Boxplot: Income",
        ylab = "Income per person",
        col = "lightblue",
        horizontal = FALSE
        )
# Miles
hist(fuel2001$Miles,
     main = "Distribution of Miles",
     xlab = "Miles of Federal-aid Highway",
     col = "lightblue",
     breaks = 20
)
boxplot(fuel2001$Income,
        main = "Boxplot: Miles",
        ylab = "Miles of Federal-aid Highway",
        col = "lightblue",
        horizontal = FALSE
)

# Pop
hist(fuel2001$Pop,
     main = "Distribution of Population",
     xlab = "Population of age 16 and over",
     col = "lightblue",
     breaks = 20
)
boxplot(fuel2001$Pop,
        main = "Boxplot: Population",
        ylab = "Population of age 16 and over",
        col = "lightblue",
        horizontal = FALSE
)
# Tax
hist(fuel2001$Tax,
     main = "Distribution of Tax",
     xlab = "Tax rate for gasoline",
     col = "lightblue",
     breaks = 20
)
boxplot(fuel2001$Income,
        main = "Boxplot: Tax",
        ylab = "Tax rate for gasoline",
        col = "lightblue",
        horizontal = FALSE
)
```

```{r}
# Install and load necessary packages
#install.packages(c("corrplot", "PerformanceAnalytics", "GGally"))
library(corrplot)
library(PerformanceAnalytics)
library(GGally)
library(ggplot2)

# Correlation Matrix
cor_matrix <- cor(fuel2001)
round(cor_matrix, 3)

# Correlation Plot
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.8,
  title = "Correlation Matrix",
  col = colorRampPalette(c("blue", "white", "red"))(200)
)

# Scatterplot Matrix
pairs(fuel2001,
      main = "Scatterplot Matrix - All Variables",
      lower.panel = NULL
)

# Custom correlation panel (fixed)
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr")           # save current coordinates
  on.exit(par(usr = usr))     # restore them after the function
  par(usr = c(0, 1, 0, 1))    # set up new normalized coordinates
  r <- cor(x, y)
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if (missing(cex.cor)) cex.cor <- 0.8 / strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * abs(r) * 2)
}

# Scatterplot Matrix with Correlations
pairs(fuel2001,
      lower.panel = panel.smooth,
      upper.panel = panel.cor,
      main = "Scatterplot Matrix with Correlation"
)

```

```{r}
# FuelC vs Drivers
ggplot(fuel2001, aes(x = Drivers, y = FuelC)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink", alpha = 0.2) +
  labs(title = "Fuel Consumption vs Drivers",
       x = "Licensed Drivers",
       y = "Fuel Consumption (1000s gallons)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# FuelC vs Income
ggplot(fuel2001, aes(x = Income, y = FuelC)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink", alpha = 0.2) +
  labs(title = "Fuel Consumption vs Income",
       x = "Per Capita Income ($1000s)",
       y = "Fuel Consumption (1000s gallons)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# FuelC vs Miles
ggplot(fuel2001, aes(x = Miles, y = FuelC)) +
  geom_point(color = "orange", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink", alpha = 0.2) +
  labs(title = "Fuel Consumption vs Highway Miles",
       x = "Federal-aid Highway Miles",
       y = "Fuel Consumption (1000s gallons)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# FuelC vs Pop
ggplot(fuel2001, aes(x = Pop, y = FuelC)) +
  geom_point(color = "purple", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink", alpha = 0.2) +
  labs(title = "Fuel Consumption vs Population",
       x = "Population (16+)",
       y = "Fuel Consumption (1000s gallons)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# FuelC vs Tax
ggplot(fuel2001, aes(x = Tax, y = FuelC)) +
  geom_point(color = "firebrick", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink", alpha = 0.2) +
  labs(title = "Fuel Consumption vs Gas Tax",
       x = "Gas Tax Rate (cents/gallon)",
       y = "Fuel Consumption (1000s gallons)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

# SECTION 2: nassCDS DATA ANALYSIS

```{r}
# install.packages("DAAG")
library(DAAG)
data(nassCDS)
```

## Question 1: Seatbelt Use and Mortality

### (1). How many indiv used seatbelts

```{r}
sum(nassCDS$seatbelt == "belted", na.rm=TRUE)
```

### (2) table of seatbelt usage vs accident outcome

```{r}
table_seat <- table(nassCDS$seatbelt, nassCDS$dead)
table_seat
```

### (3) Proportion test

```{r}
test_proportion <- function(tab){
  p1 <- tab[1,2]/sum(tab[1,])
  p2 <- tab[2,2]/sum(tab[2,])
  p <- (tab[1,2]+tab[2,2])/sum(tab)
  se <- sqrt(p*(1-p)*(1/sum(tab[1,])+1/sum(tab[2,])))
  z <- (p1 - p2)/se
  pval <- 2*pnorm(-abs(z))
  return(data.frame(Z_statistic=z, P_value=pval))
}
test_proportion(table_seat)
```

### (4) Visualization

```{r}
barplot(table_seat, beside=TRUE, col=c("skyblue","salmon"),
main="Seatbelt Usage vs Accident Outcome", ylab="Count")
```

## Question 2: Age of Occupants and Accident Outcome

```{r}
tapply(nassCDS$ageOFocc, nassCDS$dead, mean, na.rm=TRUE)
tapply(nassCDS$ageOFocc, nassCDS$dead, sd, na.rm=TRUE)

boxplot(ageOFocc ~ dead, data=nassCDS, main="Occupant Age by Outcome",
xlab="Outcome", ylab="Age", col=c("orange","lightblue"))
stripchart(ageOFocc ~ dead, data=nassCDS, vertical=TRUE, add=TRUE, pch=16, col="blue")

alive <- nassCDS$ageOFocc[nassCDS$dead=="alive"]
dead  <- nassCDS$ageOFocc[nassCDS$dead=="dead"]
t.test(alive, dead, var.equal=FALSE)
```

### (1) mean and standard deviation

```{r}
# Load the nassCDS dataset
data(nassCDS, package = "DAAG")

# Check the structure of the dataset
str(nassCDS)

# Look at the variables we need
head(nassCDS[, c("dead", "ageOFocc")])
```

```{r}
# mean and standard deviation of age by accident outcome
age_stats <- nassCDS %>%
  group_by(dead) %>%
  summarise(
    mean_age = mean(ageOFocc, na.rm = TRUE),
    sd_age = sd(ageOFocc, na.rm = TRUE),
    n = n()
  )

print(age_stats)
```

### (2) Visualize the distribution of the occupants’ age by accident outcome with data points

```{r}
# Create boxplot with data points
ggplot(nassCDS, aes(x = dead, y = ageOFocc, fill = dead)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  labs(title = "Distribution of Occupant Age by Accident Outcome",
       x = "Accident Outcome",
       y = "Age of Occupant") +
  theme_minimal() +
  scale_fill_manual(values = c("alive" = "lightblue", "dead" = "lightcoral"))
```

The boxplot shows the distribution of ages for both groups (alive and dead) with individual data points overlaid. This visualization helps compare the central tendency, spread, and potential outliers in age distribution between survivors and fatalities.

### (3) 95% confidence interval for the mean difference of the age of occupant using t distribution.

```{r}
# Calculate 95% confidence interval for mean age difference using t-distribution

# Extract ages for each group
age_alive <- nassCDS$ageOFocc[nassCDS$dead == "alive"]
age_dead <- nassCDS$ageOFocc[nassCDS$dead == "dead"]

# Perform t-test to get confidence interval
t_test_result <- t.test(age_dead, age_alive, conf.level = 0.95)

# Extract the confidence interval
ci_lower <- t_test_result$conf.int[1]
ci_upper <- t_test_result$conf.int[2]
mean_diff <- t_test_result$estimate[1] - t_test_result$estimate[2]

cat("Mean age difference (dead - alive):", round(mean_diff, 2), "\n")
cat("95% Confidence Interval: [", round(ci_lower, 2), ",", round(ci_upper, 2), "]\n")
cat("t-statistic:", round(t_test_result$statistic, 2), "\n")
cat("p-value:", format.pval(t_test_result$p.value), "\n")
```

Description:

The 95% confidence interval for the mean difference in age between occupants who died and those who survived is **\[lower bound, upper bound\]**. This means we are 95% confident that the true mean age difference in the population falls within this interval.

## Question 3: Sex, Age, and Safety Devices

### (1)

```{r}
boxplot(ageOFocc ~ sex, data=nassCDS, main="Occupant Age by Sex", col=c("pink","lightblue"))
```

### (2)

```{r}
sum(nassCDS$ageOFocc > 50 & nassCDS$dead=="alive", na.rm=TRUE)
```

### (3)

```{r}
nassCDS$AgeOFocc_class <- ifelse(nassCDS$ageOFocc > 50, 1, 0)
```

### (4-7)

```{r}
nassCDS_o50 <- nassCDS %>%
filter(ageOFocc > 50) %>%
select(dead, airbag, weight, injSeverity) %>%
na.omit()

dim(nassCDS_o50)

barplot(table(nassCDS_o50$dead, nassCDS_o50$airbag),
beside=TRUE, col=c("skyblue","tomato"),
main="Airbag vs Outcome (Age > 50)", ylab="Count")

barplot(table(nassCDS_o50$injSeverity, nassCDS_o50$airbag),
beside=TRUE, col=c("lightgreen","pink"),
main="Airbag vs Injury Severity (Age > 50)", ylab="Count")
```

## Question 4: Summary Function for Fatal Accidents

```{r}
nass_analysis <- function(data){
  d <- subset(data, dead=="dead")
  pct_dead <- 100 * nrow(d) / nrow(data)
  pct_male <- 100 * sum(d$sex=="male", na.rm=TRUE)/nrow(d)
  pct_female <- 100 * sum(d$sex=="female", na.rm=TRUE)/nrow(d)
  mode_sev <- names(sort(table(d$injSeverity), decreasing=TRUE))[1]
  age_min <- min(d$ageOFocc, na.rm=TRUE)
  age_max <- max(d$ageOFocc, na.rm=TRUE)
  table_out <- data.frame(pct_dead, pct_male, pct_female, mode_sev, age_min, age_max)
  hist(d$injSeverity, main="Injury Severity among Deaths", xlab="Injury Severity", col="lightblue")
  return(table_out)
}
nass_analysis(nassCDS)
```

### (4.1) which of the accident outcome is “dead”?

```{r}
fatal_cases <- nassCDS %>% filter(dead == "dead")

# Check the result
cat("Total observations in original dataset:", nrow(nassCDS), "\n")
cat("Fatal cases found:", nrow(fatal_cases), "\n")
cat("Percentage of fatal cases:", round(nrow(fatal_cases)/nrow(nassCDS)*100, 2), "%\n")

# Preview the fatal cases
head(fatal_cases)
```

### (4.2) The percentage of deaths

```{r}
# 4.2 Calculate percentage of deaths out of the overall number of observations
total_observations <- nrow(nassCDS)
fatal_count <- nrow(fatal_cases)
death_percentage <- (fatal_count / total_observations) * 100

cat("Overall death statistics:\n")
cat("Total observations:", total_observations, "\n")
cat("Number of deaths:", fatal_count, "\n")
cat("Percentage of deaths:", round(death_percentage, 2), "%\n")
```

### (4.3) Calculate the percentages of females and males among the occupants who died in the accident

```{r}
# 4.3 Calculate the percentages of females and males among the occupants who died
gender_breakdown <- fatal_cases %>%
  count(sex) %>%
  mutate(percentage = (n / fatal_count) * 100)

print("Gender distribution among fatalities:")
print(gender_breakdown)

# Extract specific percentages
female_pct <- gender_breakdown$percentage[gender_breakdown$sex == "f"]
male_pct <- gender_breakdown$percentage[gender_breakdown$sex == "m"]

cat("\nDetailed breakdown:\n")
cat("Female fatalities:", ifelse(length(female_pct) > 0, round(female_pct, 2), 0), "%\n")
cat("Male fatalities:", ifelse(length(male_pct) > 0, round(male_pct, 2), 0), "%\n")
```

### (4.4) Find most frequent injury severity

```{r}
# 4 Show the most frequent severity of their injuries
severity_analysis <- fatal_cases %>%
  count(injSeverity) %>%
  arrange(desc(n))  # Sort by frequency in descending order

print("Injury severity frequency among fatalities:")
print(severity_analysis)

most_frequent_severity <- severity_analysis$injSeverity[1]
most_frequent_count <- severity_analysis$n[1]

cat("\nMost frequent injury severity:", most_frequent_severity, "\n")
cat("Number of cases with this severity:", most_frequent_count, "\n")
cat("This represents", round(most_frequent_count/fatal_count*100, 2), "% of all fatalities\n")
```

This identifies the most common injury severity level among those who died. We count occurrences of each injury severity level and sort them to find the most frequent one. This reveals what injury severity is most associated with fatal outcomes.

### (4.5) Calculate the minimum and maximum age of the occupant (the variable ageOFocc).

```{r}
# 4.5 Calculate the minimum and maximum age of the occupant
min_age <- min(fatal_cases$ageOFocc, na.rm = TRUE)
max_age <- max(fatal_cases$ageOFocc, na.rm = TRUE)

cat("Min age:", min_age, "years\n")
cat("Max age:", max_age, "years\n")
cat("Age range:", max_age - min_age, "years\n")
```

### (4.6) Produce a histogram with the severity of injuries on the x axis and the occupant’s age on the y axis

```{r}
# histogram with severity of injuries on x-axis and occupant's age on y-axis
ggplot(fatal_cases, aes(x = factor(injSeverity), y = ageOFocc)) +
  geom_boxplot(aes(fill = factor(injSeverity)), alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1, color = "darkred") +
  labs(title = "Relationship Between Injury Severity and Occupant Age in Fatal Accidents",
       subtitle = paste("Analysis of", fatal_count, "fatal cases"),
       x = "Injury Severity Level",
       y = "Occupant Age (years)",
       caption = "Each point represents one fatal case") +
  theme_minimal() +
  theme(legend.position = "none") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "blue")
```

### (4.7) Complete SINGLE-for-all function

```{r}
# 4.7 Complete function that returns both numerical and graphical outputs
analyze_fatal_accidents <- function(nassCDS) {
  # Filter fatal cases
  fatal_cases <- nassCDS %>% filter(dead == "dead")
  
  # Calculate death percentage
  total_obs <- nrow(nassCDS)
  fatal_count <- nrow(fatal_cases)
  death_pct <- (fatal_count / total_obs) * 100
  
  # Gender percentages
  gender_pct <- fatal_cases %>%
    count(sex) %>%
    mutate(percentage = (n / fatal_count) * 100)
  
  # Most frequent injury severity
  most_severe <- fatal_cases %>%
    count(injSeverity) %>%
    arrange(desc(n)) %>%
    slice(1)
  
  # Age range
  min_age <- min(fatal_cases$ageOFocc, na.rm = TRUE)
  max_age <- max(fatal_cases$ageOFocc, na.rm = TRUE)
  
  # Create numerical output table
  numerical_output <- data.frame(
    Metric = c("Death Percentage", "Female Fatalities %", "Male Fatalities %", 
               "Most Frequent Severity", "Min Age", "Max Age"),
    Value = c(round(death_pct, 2),
              round(gender_pct$percentage[gender_pct$sex == "f"], 2),
              round(gender_pct$percentage[gender_pct$sex == "m"], 2),
              most_severe$injSeverity,
              round(min_age, 1),
              round(max_age, 1))
  )
  
  # Create plot
  graphical_output <- ggplot(fatal_cases, aes(x = factor(injSeverity), y = ageOFocc)) +
    geom_boxplot(aes(fill = factor(injSeverity)), alpha = 0.7) +
    labs(title = "Injury Severity vs Age in Fatal Accidents",
         x = "Injury Severity", y = "Occupant Age") +
    theme_minimal() +
    theme(legend.position = "none")
  
  return(list(numerical = numerical_output, plot = graphical_output))
}

# Execute the complete function
results <- analyze_fatal_accidents(nassCDS)
```

```{r}
print(results$numerical)
```

```{r}
print(results$plot)
```

## Question 5

```{r}
seatbelt_users <- nassCDS[nassCDS$seatbelt=="belted", ]
nrow(seatbelt_users)
table(seatbelt_users$dead)
table(seatbelt_users$dead, seatbelt_users$occRole)

seatbelt_users <- seatbelt_users[order(seatbelt_users$injSeverity, seatbelt_users$ageOFocc), ]
head(seatbelt_users[order(-seatbelt_users$weight), ], 25)
```

### (5.1) Create a new data frame which contains only occupants who used seatbelt.

```{r}
seatbelt_users <- nassCDS %>% filter(seatbelt == "belted")

# Check the result
head(seatbelt_users)
```

### (5.2) How many occupants used seatbelt ?

```{r}
total_seatbelt_users <- nrow(seatbelt_users)

cat("Occupants who used seatbelt:", total_seatbelt_users, "\n")
```

### (5.3) Among the individuals who used seatbelt, how many died and how many survived the accident ?

```{r}
# Among seatbelt users, how many died and how many survived?
survival_breakdown <- seatbelt_users %>%
  count(dead) %>%
  mutate(percentage = (n / total_seatbelt_users) * 100)

# Calculate fatality rate
fatalities <- survival_breakdown$n[survival_breakdown$dead == "dead"]
survivors <- survival_breakdown$n[survival_breakdown$dead == "alive"]

cat(
  "Seatbelt users who survived:", 
  survivors,
  "(", round(survivors/total_seatbelt_users*100, 2), 
  "%)\n"
)
cat(
  "Seatbelt users who died:", 
  fatalities, 
  "(", round(fatalities/total_seatbelt_users*100, 2), 
  "%)\n"
)
```

### (4) Among the individuals who used seatbelt, how many were drivers among the individuals who died and how many were passengers among the individuals who survived the accident (use the variable occRole to identify drivers/passengers) ?

```{r}
# Analyze drivers vs passengers among different outcomes
role_analysis <- seatbelt_users %>%
  group_by(dead, occRole) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(dead) %>%
  mutate(percentage = (count / sum(count)) * 100)

cat("Role distribution among seatbelt users by outcome:\n")
print(role_analysis)

# Specific answers for the question
drivers_died <- seatbelt_users %>%
  filter(dead == "dead" & occRole == "driver") %>%
  nrow()

passengers_survived <- seatbelt_users %>%
  filter(dead == "alive" & occRole == "passenger") %>%
  nrow()

cat("\nSpecific counts requested:\n")
cat("Drivers among seatbelt users who died:", drivers_died, "\n")
cat("Passengers among seatbelt users who survived:", passengers_survived, "\n")

# Additional insight: compare role distribution
cat("\nAdditional insight - Role distribution:\n")
role_summary <- seatbelt_users %>% count(occRole)
print(role_summary)
```

### (5.5) Sort by injury severity and age

```{r}
# Sort the data frame according to injury severity and occupant age
seatbelt_sorted <- seatbelt_users %>%
  arrange(desc(injSeverity), desc(ageOFocc))

cat("Data frame sorted by injury severity (descending) and age (descending)\n")
cat("First 10 rows of sorted data:\n")
print(seatbelt_sorted[1:10, c("injSeverity", "ageOFocc", "dead", "occRole", "weight")])

# Alternative: ascending order for age
seatbelt_sorted_alt <- seatbelt_users %>%
  arrange(desc(injSeverity), ageOFocc)

cat("\nAlternative: Sorted by injury severity (desc) and age (asc)\n")
cat("First 10 rows:\n")
print(seatbelt_sorted_alt[1:10, c("injSeverity", "ageOFocc", "dead", "occRole", "weight")])

# Show range of injury severity
cat("\nInjury severity range in seatbelt users:\n")
cat("Min severity:", min(seatbelt_users$injSeverity, na.rm = TRUE), "\n")
cat("Max severity:", max(seatbelt_users$injSeverity, na.rm = TRUE), "\n")
cat("Mean severity:", round(mean(seatbelt_users$injSeverity, na.rm = TRUE), 2), "\n")
```

### (5.6) Top 25 occupants by weight

```{r}
# Print the 25 occupants with the highest weight
top_25_weight <- seatbelt_users %>%
  arrange(desc(weight)) %>%
  head(25)

print(top_25_weight[, c("weight", "ageOFocc", "dead", "injSeverity", "occRole", "sex")])
```

## Question 6: Age, Seatbelt, and Airbag Relationship

```{r}
# Plot
ggplot(nassCDS, aes(x = factor(injSeverity), y = ageOFocc, fill = injSeverity)) +
  geom_boxplot() +
  scale_fill_gradient(low = "darkblue", high = "skyblue") +
  facet_grid(seatbelt ~ airbag) +
  labs(
    x = "injSeverity",
    y = "ageOFocc",
    title = NULL,
    fill = "injSeverity"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "lightgray", color = NA),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )
```
